---
title: "Data Wrangling - 7b. Safe Iteration"
author: "Kieran Healy"
date: "`r Sys.Date()`"
output: kjhslides::kjh_slides_reader
editor_options: 
  chunk_output_type: console
---


```{r packages, include=FALSE}
library(flipbookr)
library(here)
library(tidyverse)
library(kjhslides)
```


```{r setup, include=FALSE}

kjh_register_tenso()
kjh_set_knitr_opts()
kjh_set_slide_theme()
kjh_set_xaringan_opts()

```


class: center middle main-title section-title-1

# .kjh-green[Safely] iterating with .kjh-yellow[purrr] and .kjh-yellow[map]  

.class-info[

**Data Wrangling: Session 7b**

.light[Kieran Healy<br>
Statistical Horizons, April 2022]

]

---

layout: true
class: title title-1

---

# Load the packages, as always

.SMALL[
```{r 07b-iterating-safely-2, message = TRUE}
library(here)      # manage file paths
library(socviz)    # data and some useful functions
library(tidyverse) # your friend and mine
```
]

---

# Additional libraries

```{r 07b-iterating-safely-3}
library(survey)
library(srvyr)
library(broom)
library(gssr) # https://kjhealy.github.io/gssr
```


---

# The complete GSS

```{r 07b-iterating-safely-4}
data(gss_all)

gss_all


```

---

# Set up our analysis

```{r 07b-iterating-safely-5}
cont_vars <- c("year", "id", "ballot", "age")
cat_vars <- c("race", "sex", "fefam")
wt_vars <- c("vpsu",
             "vstrat",
             "oversamp",
             "formwt",              # weight to deal with experimental randomization
             "wtssall",             # main weight variable
             "sampcode",            # sampling error code
             "sample")              # sampling frame and method
my_vars <- c(cont_vars, cat_vars, wt_vars)

```

---

# Clean the labeled variables

```{r 07b-iterating-safely-6}
gss_df <- gss_all %>%
  filter(year > 1974 & year < 2021) %>% 
  select(all_of(my_vars)) %>% 
  mutate(across(everything(), haven::zap_missing), # Convert labeled missing to regular NA
         across(all_of(wt_vars), as.numeric),
         across(all_of(cat_vars), as_factor), 
         across(all_of(cat_vars), fct_relabel, tolower),
         across(all_of(cat_vars), fct_relabel, tools::toTitleCase),
         compwt = oversamp * formwt * wtssall)

```

---

# Working dataset

```{r 07b-iterating-safely-7}
gss_df
```

---


# The `fefam` question

```{r 07b-iterating-safely-8}
gss_df %>% 
  count(fefam) 
```

---

# Recoding

```{r 07b-iterating-safely-9}
gss_df <- gss_df %>% 
  mutate(fefam_d = forcats::fct_recode(fefam,
                                  Agree = "Strongly Agree",
                                  Disagree = "Strongly Disagree"),
    fefam_n = recode(fefam_d, "Agree" = 1, "Disagree" = 0))

# factor version
gss_df %>% 
  count(fefam_d) 

# numeric version, 1 is "Agree"
gss_df %>% 
  count(fefam_n) 

```

---

# Unweighted model

```{r 07b-iterating-safely-10}
out_all <- glm(fefam_n ~ age + sex + race, 
              data = gss_df, 
              family="binomial", 
              na.action = na.omit)

summary(out_all)

```

---

# Tidied output

```{r 07b-iterating-safely-11}
tidy(out_all)
```

---

# .kjh-green[**`group_map()`**] and .kjh-green[**`possibly()`**]

- Model each year

```{r 07b-iterating-safely-12}
out_yr <- gss_df %>% 
  group_by(year) %>% 
  group_map_dfr(possibly(~ tidy(glm(fefam_n ~ age + sex + race, 
                       data = .x, 
                       family = "binomial", 
                       na.action = na.omit), 
                       conf.int = TRUE), 
                     otherwise = NULL))

out_yr

```

---

# .kjh-green[**`group_map()`**] and .kjh-green[**`possibly()`**]

```{r 07b-iterating-safely-13, echo = TRUE, eval = FALSE}
possibly(~ tidy(glm(...)), otherwise = NULL)
```

---

# .kjh-green[**`group_map()`**] and .kjh-green[**`possibly()`**]

```{r 07b-iterating-safely-14, fig.width=12, fig.height=5}
out_yr %>% 
  filter(term == "sexFemale") %>% 
  ggplot(mapping = aes(x = year, y = estimate,
                       ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_line() + 
  geom_pointrange()
```

---

# Survey-weighted estimates

```{r 07b-iterating-safely-15}
options(survey.lonely.psu = "adjust")
options(na.action="na.pass")

gss_svy <- gss_df %>%
  filter(year > 1974) %>%  
  mutate(stratvar = interaction(year, vstrat)) %>%
  as_survey_design(id = vpsu,
                     strata = stratvar,
                     weights = compwt,
                     nest = TRUE)
gss_svy

```

---


# Survey-weighted estimates

```{r 07b-iterating-safely-16}
gss_svy %>%
  drop_na(fefam_d) %>% 
  group_by(year, sex, race, fefam_d) %>%
  summarize(prop = survey_mean(na.rm = TRUE, 
                               vartype = "ci"))

```

---

# Survey-weighted estimates


```{r 07b-iterating-safely-17}
out_svy_all <- svyglm(fefam_n ~ age + sex + race, 
                  design = gss_svy, 
                  family = quasibinomial(),
                  na.action = na.omit)

tidy(out_svy_all)
```

---

# Survey-weighted estimates

```{r 07b-iterating-safely-18}
out_svy_yrs <- gss_svy %>% 
  group_by(year) %>% 
  group_map_dfr(possibly(~ tidy(svyglm(fefam_n ~ age + sex + race, 
                       design = .x, 
                       family = quasibinomial(),
                       na.action = na.omit),
                       conf.int = TRUE), 
                     otherwise = NULL))

out_svy_yrs

```

---

# Survey-weighted estimates

```{r 07b-iterating-safely-19, fig.height=5, fig.width=12}
out_svy_yrs %>% 
  filter(term == "sexFemale") %>% 
  ggplot(mapping = aes(x = year, 
                       y = estimate,
                       ymin = conf.low,
                       ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_line() + 
  geom_pointrange()
```


