---
title: "Template Slides"
author: "Kieran Healy"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    chakra: "libs/remark-latest.min.js"
    css: ["default", "css/kjh-slides.css", "css/kjh-inferno-fonts.css", "css/animate.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```

```{r packages-data, include=FALSE}
library(flipbookr)
library(here)
library(tidyverse)
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
xaringanExtra::use_animate_css()
xaringanExtra::use_animate_all("fade")
xaringanExtra::use_clipboard()
```

class: center middle main-title section-title-1

# A Title Slide

.class-info[

**Session 1**

.light[Kieran Healy<br>
Statistical Horizons, April 2021
]

]

---

layout: true
class: title title-1

---

# A Bullet Point Slide

&nbsp;

--

.box-inv-1.medium[Blah]

--

.box-inv-1.medium[Blah]

--

.box-inv-1.medium.sp-after[Blah]

---

name: bigtext
layout: false
name: program-theories
class: center middle section-title section-title-1

# Most Data Analysis<br />is Cleaning &<br />Recoding

---

layout: true
class: title title-1

```{r load_penguins, include = FALSE}

library(palmerpenguins)

```

`r chunk_reveal("penguin_tabs", widths = c(35,65), title = "## Breaks")`

```{r penguin_tabs, include = FALSE}
penguins %>% 
  filter(year != 2007) %>% 
  group_by(species) %>% 
  summarize(across(starts_with("bill"), 
                   list(mean = mean, 
                        sd = sd),
                   .names = "{.fn}_{.col}",
                   na.rm = TRUE)) 
```

---

```{r my_cars, include = FALSE}
cars %>%
  filter(speed > 4) %>%
  ggplot() +
  aes(x = speed) +
  aes(y = dist) + #BREAK
  geom_point(
    alpha = .8, 
    color = "blue" 
    ) + #BREAK
  aes(size = speed) #BREAK
```

`r chunk_reveal("my_cars", break_type = "auto", title = "# Chunk Breaks")`


