---
title: "Data Wrangling with R and the tidyverse"
description: |
  A Short Course using R and the Tidyverse suite of tools.
author:
- name: Kieran Healy
  url: https://kieranhealy.org
date: "`r Sys.Date()`"
output: distill::distill_article
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r}
library(tidyverse)
library(socviz)
```

# Selecting, Mutating and Summarizing

The `ukelection2019` package has a tibble called `ukvote2019` that contains the results of the UK's 2019 general election, by candidate. 

```{r}
library(ukelection2019)
ukvote2019
```

The data are (accidentally) grouped by candidate, so let's ungroup it first.

```{r}
ukvote2019 <- ukvote2019 %>% 
  ungroup()
```

## Some things to find

- What were the results in the Dover constituency?

```{r}

```


- How many constituencies (US translation: districts) are there?

- How many candidates did each party run in the election as a whole?

- Show the number of candidates fielded by party from highest to lowest.

- Calculate the percentage of all candidates fielded by each party

- Find the five worst-performing candidates overall, as measured by number of votes.

- Find the five largest constituencies in the country as measured by votes cast

- Find the total number of votes cast nationally for the Labour and Conservative parties.

- Find the ten largest vote-winning candidates in the country, as measured by percentage of vote share.




